Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-23 paper.tex
Yan's updates

% Paragraph 2, what do we plot, specifically?
In this section we investigate the patterns of divergence between
the parents of recombination nodes.
We focus on the Long ARG because  it covers time
periods where recombination is known to have had a significant
impact, and includes
[HOW MANY?]
samples designated to [HOW MANY] Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis of Pango recombinant lineages in the Long and Wide ARGs).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e., the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, we exclude recombination nodes
that are ancestral to only one sample from this analysis.
Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all recombinants that
are ancestral to more than one sample, and
Figure~\ref{fig:recomb_mrcas}B plots a subset which pass some further
reliability filters. Specifically, only recombination nodes that
[FILL ME IN].
This reduces the number of recombination nodes
to 347 and breakpoints to 358.
Note that each plotted point
represents a breakpoint, so that a recombination node with more than one
breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
It is important to note here that some of the time estimates in these figures
come with a substantial degree of uncertainly,
since nodes other than samples are dated somewhat crudely in the current version
of \texttt{sc2ts} (Section~\ref{sec:node_dating}).

@@

In this section we focus on the detailed genealogy of samples that have been
previously identified as recombinants, i.e. given a Pango designation that
starts with an ``X''. Since the Wide ARG is restricted to data collected prior
to mid-2021, it contains samples from only three Pango-designated recombinant
lineages: XA, XB, and XC. To investigate the diversity of genealogies
associated with recombinants, we therefore devote more analysis to patterns in
the Long ARG. As in previous sections, we exclude singleton recombinants where
possible.

\subsubsection{Wide ARG}
%% Quick overview of the wide ARG.
In the Wide ARG, 44
samples are designated as XA by both Nextclade and GISAID, 237 are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
designated as XA, XB, or XC by any method have one or more recombination nodes
in their ancestry.

%% Textual description of the 3 recombinant lineages present in the wide ARG
The simplest Pango recombinant genealogy is XA: all XA samples trace back to a
unique originating recombination node. This is the product of a recombination
between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
EPI\_ISL: 1012804) which contributed the majority of the genome from the start
to a maximum position of 22227bp, and an unknown (inserted) node with imputed
Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
recombination node has five immediate children: four sample leaves (EPI\_ISLs:
989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
ancestor of all other XA samples in the dataset.

For XB, all samples trace back to a originating node which is the product of a
recombination between a B.1 sample (specifically the strain
England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
the genome from the start up to a maximum position of 23604bp), and two other
UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
27389bp) and B.1.36.8 (the remaining fragment of the genome). As described
later, in the long ARG the equivalent recombination node has only 2 parents,
with no involvement of  B.1.36.8, and it is possible that the third parent in
the wide ARG is artefactual. Additional complications arise because additional
non-X-designated samples such as B.1.634 also descend from this recombination
(also true in the long ARG, see below), and there are also a small number of
further recombination nodes nested within the XB ``clade''. However, all but
one of these nested recombinations are ancestral to negligible fractions
of the designated XB samples. The exception accounts for about 17\% of the
designated XB nodes, and involves a recombination between descendants of the
originating XB recombination. More specifically strain
USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination
between an XB grandparent and its XB grandchild. This could well be
artefactual.

For XC, the Wide ARG identifies more than one originating recombination node
(see "multiple origins" below). However, as none of the samples designated as
XC by Nextclade are designated as XC by GISAID, these patterns could be due to
mistaken labelling, hence we do not examine XC further.

@@

One of the simplest examples is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
Here, a subgraph of the long ARG has been plotted, where paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset.
Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
non-sample nodes: i.e. inserted nodes such as UPGMA, recombination, mutation collapse, and
reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
that the subgraph is not extended to show additional descendants of recombinant parents.

It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a single originating recombination node,
whose genome is a composite of a B.1.177.18 lineage on the left of the genome (up to a maximum
position of ${\sim}22227$ basepairs) and a B.1.1.7 lineage for the remainder of the genome.

@@

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. 15 Pango X lineages are inferred to have multiple
recombinant origins, such that their samples trace to $>$1 most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e. there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend (but in XJ and XU there are too few Pango X samples to decide on a ``main'' clade:
see Supplementary Table~\ref{tab:pango-recombinants}).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
the 4 samples labelled XD by Nextclade in the Long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see Supplementary
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).

A similar pattern is seen for several of the origins inferred for the Pango
XN lineage (originally defined on the basis of sequences from the UK). After removing
singleton recombinants and the descendants of the erroneously recombinant
Denmark/DCGC-281594/2021 strain (which, however, includes 53 XN-designated samples)
the Long ARG contains four independent origins for Nextclade-designated XN
samples (Supplementary Table~\ref{tab:multiple_origins_table}). The oldest
of these four origins has 3 mutations above the sample that caused the
recombination to be inferred. It therefore seems likely that we have not managed
to sample an XN strain that is very close to the true recombinant. This may
lead to \texttt{sc2ts} mistakenly inferring multiple recombination events when there

If the only samples available are rather genetically distant from an originating
recombinant, we might expect sc2ts to infer multiple similar recombination events
(rather than a single event with extra reversions)

a recombinant outbreak
has taken a ``true'' XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
is presumably because it has changed too much from the nearest in the main XE cluster - need to discuss the
weighting of 3 mutations for one recombination here. NB: the number
of nucleotide differences between the closest XE in the two clusters is 11: see the
Pango-recombinant-analysis workbook]

\subsubsection{Nested recombinant origins}
As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked with asterisks in Supplementary Table~\ref{tab:pango-recombinants}; the most
extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
The Logn ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the Wide ARG (but as discussed previously,
in the Wide ARG, XB exhibits multiple nested recombinations, some with many XB
descendants).

@@

@@

\section{Data Availability}
The source code for \texttt{sc2ts} is
is available on GitHub
\url{https://github.com/jeromekelleher/sc2ts/}.
The Jupyter notebooks and code used to produce the results described here are also
available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.

@@

\begin{sidewaystable}
% \begin{footnotesize}
\centering
\begin{tabular}{p{1cm}p{1.2cm}p{4.2cm}lll}
\toprule
Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: additional descendants \\
\midrule
\bfseries XA & 1 & \textbf{5} & B.1.1.7 + B.1.177 & B.1.1.7 + B.1.177.18 &  \\
\bfseries XF & 1 & \textbf{2} & B.1.617.2* + BA.1* & AY.4 + BA.1 &  \\
\bfseries XG & 1 & \textbf{32} & BA.1* + BA.2* & BA.1.17 + BA.2 & XAB: 1/48 \\
\bfseries XH & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XAF: 34/35, B.1.1.529: 2, XE: 3/163 \\
\bfseries XK & 1 & \textbf{3} & BA.1* + BA.2* & BA.1.1.1 + BA.2 &  \\
\bfseries XL & 1 & \textbf{10} & BA.1* + BA.2* & BA.1.17.2 + BA.2 & XAB: 1/48, XU: 1/3 \\
\bfseries XR & 1 & \textbf{8} & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XQ: 9/12, XAB: 2/48 \\
\bfseries XS & 1 & \textbf{4} & B.1.617.2* + BA.1.1* & AY.36 + BA.1.1 &  \\
\bfseries XT & 1 & \textbf{1} & BA.1* + BA.2* & BA.2.23 + Unknown & BA.2.23: 1 \\
\bfseries XV & 1 & \textbf{1} & BA.1* + BA.2* & BA.1.1 + BA.2 + Unknown & BA.2: 8 \\
\bfseries XW & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.1.15 + BA.2 & XN: 2/13 \\
\bfseries XY & 1 & \textbf{6} & BA.1* + BA.2* & BA.1.1 + BA.2 & XAF: 1/35 \\
\bfseries XAA & 1 & \textbf{8} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bfseries XAC & 1 & \textbf{27} & BA.1* + BA.2* & BA.1.17.2 + BA.2.3 &  \\
\bfseries XAE & 1 & \textbf{9} & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAG & 1 & \textbf{17} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAA: 8/8, XU: 1/3, XQ: 2/12 \\
\bfseries XB & 2 & 57 (\textbf{57}, 1)* & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
\bfseries XC & 2 & 3 (1, \textbf{2}) & AY.29 + B.1.1.7 & AY.103 + B.1.1.7 &  \\
\bfseries XD & 2 & 4 (1, \textbf{3}) & B.1.617.2* + BA.1* & AY.4 + BA.1.15 &  \\
\bfseries XJ & 2 & 2 (1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XAD & 2 & 5 (1, \textbf{4}) & BA.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XAF & 2 & 35 (\textbf{34}, 1) & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XH: 11/11, B.1.1.529: 2, XE: 3/163 \\
\bfseries XAH & 2 & 12 (\textbf{10}, 2) & BA.1* + BA.2* & BA.1 + BA.2.10 & XZ: 61/66, XAD: 1/5 \\
\bfseries XQ & 3 & 12 (2, \textbf{9}, 1) & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XR: 8/8, XAB: 2/48 \\
\bfseries XU & 3 & 3 (1, 1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XM & 4 & 47 (2, 1, \textbf{40}, 4) & BA.1.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XN & 4 & 13 (2, 2, \textbf{7}, 2) & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)* & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)* & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
\bfseries XZ & 5 & 66 (1, 1, \textbf{61}, 2, 1) & BA.1* + BA.2* & BA.1 + BA.2.10 & XAH: 10/12, XAD: 1/5 \\
\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)* & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bottomrule
\end{tabular}
\caption{\label{tab:pango-recombinants}
Summary of the Pango X-lineages in the long ARG (excluding XP and XAK whose samples are entirely filtered out, see text). In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold).
Sc2ts inferred parents for the main clade are based on imputed NextClade Pango lineages. Official parents
are taken from Pango designation alias key (TODO: explain the asterisks). Other inferred samples within the main clade are listed, grouped by NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
}
\end{sidewaystable}

\begin{table}
\small
\centering
\begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
\toprule
Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
\midrule
2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
\bottomrule
\end{tabular}
\caption{\label{tab:multiple_origins_table}
Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
(Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
The 13 XN samples are grouped into rows based on their most recent ancestral recombination node;
all the descendant samples of the four recombination nodes are designated XN by Nextclade.
Details are also provided for the ``causal strain'' that led to the recombination
being proposed by \texttt{sc2ts}. When undersampling causes three or more mutations to lie
immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple
independent origins rather than one. Note that a large number of XN samples were previously
removed by filtering.
}
\end{table}

>>>>

% Paragraph 2, what do we plot, specifically?
In this section we investigate the patterns of divergence between
the parents of recombination nodes.
We focus on the Long ARG because it covers time
periods where recombination is known to have occurred
a large number of times, for example it contains samples from over
30 Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis of Pango recombinant lineages in the Long and Wide ARGs).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e. the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, in these plots we
exclude ``singleton'' recombination nodes (those ancestral to only one sample).
Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all nonsingleton recombinants,
and Figure~\ref{fig:recomb_mrcas}B plots a subset of potentially more
reliable breakpoints that occur in nodes
ancestral to 10 or more samples (note that this is one of the official criteria
used when proposing Pango recombinant lineages [CITE]).
% that we deem to be more reliable by the measure of ``HMM consistency''. 
% We define an  in that when we compare the LS HMM results when run from
% 5' to 3' versus 3' to 5', it gives the
% same mutations patterns and sequence-identical 5' parents as when run 3' to 5'
This reduces the number of recombination nodes
to XXX and breakpoints to YYY.
Note that each plotted point
represents a breakpoint, so that a recombination node with more than one
breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
It is important to note here that some of the time estimates in these figures
come with a substantial degree of uncertainly,
since nodes other than samples are dated somewhat crudely in the current version
of \texttt{sc2ts} (Section~\ref{sec:node_dating}). Nevertheless, the histograms
clearly show that \texttt{sc2ts} can identify recombination between lineages
that are only a matter of weeks apart.

@@

In this section we focus on the detailed genealogy of samples that have been
previously identified as recombinants, i.e. given a Pango designation that
starts with an ``X''. As the Wide ARG was only sampled until mid 2021,
the majority of these Pango X lineages are found only in the Long ARG.

@@

One of the simplest examples is the case of XA, corresponding to group A
of \citet{Jackson2021-ik} as discussed above. Figure~\ref{fig:pango-simple-origin-graph}A
shows the exact relationships inferred by \texttt{sc2ts} as a subgraph of the Long ARG. 
Here, paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in the ARG.
Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
non-sample nodes: i.e. inserted nodes such as UPGMA, recombination, mutation collapse, and
reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
that the subgraph is not extended to show additional descendants of recombinant parents.

It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a
single originating recombination node, whose genome is a composite of a B.1.177.18 lineage
on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
the maximum genomic position for the recombination breakpoint, here at 22227 basepairs
(corresponding to a breakpoint strictly less than 22228, see Table~\ref{tab:jackson})

@@

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. 15 Pango X lineages are inferred to have multiple
recombinant origins, such that their samples trace to $>$1 most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e. there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend, and which agrees with the offical Pango designated parent lineages
(see Supplementary Table~\ref{tab:pango-recombinants}, but note that
in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
the 4 samples labelled XD by Nextclade in the Long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619, not shown) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see Supplementary
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).
Investigation of other multiple-origin examples reveals somewhat similar patterns,
suggesting that most of the simple multiple origin examples are due to Nextclade mislabelling.

\subsubsection{Nested recombinant origins}
As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked with asterisks in Supplementary Table~\ref{tab:pango-recombinants}; the most
extensive is for XAB). The earliest example (present in both the Wide and Long ARGs) is XB
The Long ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the Wide ARG (see below).

@@

\subsubsection{Wide ARG}
%% Quick overview of the wide ARG.
Since the Wide ARG is restricted to data collected prior
to mid-2021, it contains samples from only three Pango-designated recombinant
lineages: XA, XB, and XC. Both Nextclade and GISAID designate 44
samples as XA, while 237 samples are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 Nextclade-designated samples (229 GISAID) and XC is reduced to 4
(none in the GISAID designations). We correctly identify that all samples
designated as XA, XB, or XC by any method have one or more recombination nodes
in their ancestry.

%% Textual description of the 3 recombinant lineages present in the wide ARG
As in the Long ARG, all XA samples in the wide ARG trace back to a
unique originating recombination node. This is the product of a recombination
between a B.1.177.18 sample (specifically the strain Wales/ALDP-115BF41/2021,
EPI\_ISL: 1012804) which contributed the majority of the genome from the start
to a maximum position of 22227, and an unknown (inserted) node with imputed
Pango lineage B.1.1.7, which contributed the remaining right hand portion. The
recombination node has five immediate children: four sample leaves (EPI\_ISLs:
989697, 1019487, 1104468 and 1122630) and a UPGMA-inserted node which is the
ancestor of all other XA samples in the dataset.

For XB, all samples trace back to an originating node which is the product of a
recombination between a B.1 sample (specifically the strain
England/CAMB-7B47D/2020, EPI\_ISL\_433960, which contributed the majority of
the genome from the start up to a maximum position of 23604bp), and two other
UPGMA nodes with imputed Pango lineages B.1.627 (up to a maximum position of
27389bp) and B.1.36.8 (the remaining fragment of the genome). Above, we noted that
in the long ARG the equivalent recombination node has only 2 parents,
with no involvement of  B.1.36.8; it is possible that the third parent in
the wide ARG is artefactual. As in the Long ARG, additional
non-X-designated samples such as B.1.634 also descend from this recombination,
and there are also a small number of nested recombination nodes. However, all but
one of these nested recombinations are unimportant, being ancestral to negligible fractions
of the Nextclade-designated XB samples. The one exception accounts for about 17\% of the
designated XB nodes, and involves a recombination between descendants of the
originating XB recombination. More specifically strain
USA/TX-HMH-MCoV-43092/2021 (EPI\_ISL 2224652) is inferred to be a recombination
between an XB grandparent and its XB grandchild. This could well be
artefactual.

For the few XC-labelled samples, the Wide ARG identifies more than one
originating recombination node. However, as none of the samples designated as
XC by Nextclade are designated as XC by GISAID, these patterns could be due to
mislabelling, and greater number of XC samples would be needed to draw reasonable
conclusions.

@@

\section{Data Availability}
The source code for \texttt{sc2ts} is 
available on GitHub
\url{https://github.com/jeromekelleher/sc2ts/}.
The Jupyter notebooks and code used to produce the results described here are also
available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.

@@

\begin{sidewaystable}
% \begin{footnotesize}
\centering
\begin{tabular}{p{1cm}p{1.2cm}p{4.2cm}lll}
\toprule
Focal Pango  & Num origins & Num focal samples (further split by origin, \textdagger=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: additional descendants \\
\midrule
\bfseries XA & 1 & \textbf{5} & B.1.1.7 + B.1.177 & B.1.1.7 + B.1.177.18 &  \\
\bfseries XF & 1 & \textbf{2} & B.1.617.2* + BA.1* & AY.4 + BA.1 &  \\
\bfseries XG & 1 & \textbf{32} & BA.1* + BA.2* & BA.1.17 + BA.2 & XAB: 1/48 \\
\bfseries XH & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XAF: 34/35, B.1.1.529: 2, XE: 3/163 \\
\bfseries XK & 1 & \textbf{3} & BA.1* + BA.2* & BA.1.1.1 + BA.2 &  \\
\bfseries XL & 1 & \textbf{10} & BA.1* + BA.2* & BA.1.17.2 + BA.2 & XAB: 1/48, XU: 1/3 \\
\bfseries XR & 1 & \textbf{8} & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XQ: 9/12, XAB: 2/48 \\
\bfseries XS & 1 & \textbf{4} & B.1.617.2* + BA.1.1* & AY.36 + BA.1.1 &  \\
\bfseries XT & 1 & \textbf{1} & BA.1* + BA.2* & BA.2.23 + Unknown & BA.2.23: 1 \\
\bfseries XV & 1 & \textbf{1} & BA.1* + BA.2* & BA.1.1 + BA.2 + Unknown & BA.2: 8 \\
\bfseries XW & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.1.15 + BA.2 & XN: 2/13 \\
\bfseries XY & 1 & \textbf{6} & BA.1* + BA.2* & BA.1.1 + BA.2 & XAF: 1/35 \\
\bfseries XAA & 1 & \textbf{8} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bfseries XAC & 1 & \textbf{27} & BA.1* + BA.2* & BA.1.17.2 + BA.2.3 &  \\
\bfseries XAE & 1 & \textbf{9} & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAG & 1 & \textbf{17} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAA: 8/8, XU: 1/3, XQ: 2/12 \\
\bfseries XB & 2 & 57 (\textbf{57}, 1)\textdagger & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
\bfseries XC & 2 & 3 (1, \textbf{2}) & AY.29 + B.1.1.7 & AY.103 + B.1.1.7 &  \\
\bfseries XD & 2 & 4 (1, \textbf{3}) & B.1.617.2* + BA.1* & AY.4 + BA.1.15 &  \\
\bfseries XJ & 2 & 2 (1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XAD & 2 & 5 (1, \textbf{4}) & BA.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XAF & 2 & 35 (\textbf{34}, 1) & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XH: 11/11, B.1.1.529: 2, XE: 3/163 \\
\bfseries XAH & 2 & 12 (\textbf{10}, 2) & BA.1* + BA.2* & BA.1 + BA.2.10 & XZ: 61/66, XAD: 1/5 \\
\bfseries XQ & 3 & 12 (2, \textbf{9}, 1) & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XR: 8/8, XAB: 2/48 \\
\bfseries XU & 3 & 3 (1, 1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XM & 4 & 47 (2, 1, \textbf{40}, 4) & BA.1.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XN & 4 & 13 (2, 2, \textbf{7}, 2) & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)\textdagger & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)\textdagger & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
\bfseries XZ & 5 & 66 (1, 1, \textbf{61}, 2, 1) & BA.1* + BA.2* & BA.1 + BA.2.10 & XAH: 10/12, XAD: 1/5 \\
\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)\textdagger & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bottomrule
\end{tabular}
\caption{\label{tab:pango-recombinants}
Summary of the Pango X-lineages in the long ARG (excluding XP and XAK whose samples are entirely filtered out, see text). In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold).
Sc2ts inferred parents for the main clade are based on Nextclade designations, imputed where necessary.
Official parents are taken from Pango designation alias key (TODO: explain the asterisks). Note that B.1.617.2 is the origin of the Delta VoC, which includes all AY.* classes. Additional descendants within the main clade are summarised giving their NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
}
\end{sidewaystable}

% \begin{table}
% \small
% \centering
% \begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
% \toprule
% Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
% \midrule
% 2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
% 2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
% 2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
% 2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
% \bottomrule
% \end{tabular}
% \caption{\label{tab:multiple_origins_table}
% Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
% (Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
% on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
% The 13 XN samples are grouped into rows based on their most recent ancestral recombination node;
% all the descendant samples of the four recombination nodes are designated XN by Nextclade.
% Details are also provided for the ``causal strain'' that led to the recombination
% being proposed by \texttt{sc2ts}. When undersampling causes three or more mutations to lie
% immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple
% independent origins rather than one. Note that a large number of XN samples were previously
% removed by filtering.
% }
% \end{table}
